Nengo: A Python tool for building large-scale functional brain models

### 2013-09-25 nengo.tex
New introduction based on feedback

Nengo is a set of software tools built around
a theoretical framework for building neural models
proposed by Eliasmith \& Anderson \cite{TODO},
and more recently, an architecture
for performing high-level cognitive tasks
in that framework proposed by Eliasmith \cite{TODO}.
Nengo consists of a Java-based simulator,
a Jython-based scripting interface
for common use cases of the simulator,
and a Java-based GUI that
uses both the underlying simulator
and the scripting interface.
This set of tools has been used
to create models of visual attention \cite{TODO},
inductive reasoning \cite{TODO},
and reinforcement learning \cite{TODO},
as well as Spaun, a cognitive agent able
to do these and several other tasks
(\cite{TODO}).
Nengo has been used by dozens of researchers
in labs at the University of Waterloo,
Johns Hopkins University, Yale, Stanford,
and others (TODO: confirm).

Here, we describe a new tool
for creating and simulating NEF-based models
implemented in Python,
which effectively merges two existing tools
into one unified package
that can be extended in
several ways that previous Nengo tools could not.

\subsection{Why reimplement Nengo?}

\begin{table}[!t]
\processtable{Nengo tools\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{Footnote?}
\end{table}

The original Java simulator was created
with the intention of being a general
neural simulator that included
the methods of the Neural Engineering Framework
(NEF; \cite{TODO}) as one of many different
methods of creating neural models.
However, since creating the simulator in 2007,
it has only been extended minimally,
but has been used extensively for its ability
to create NEF-based models.

The Jython scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
However, the Java GUI was developed at the same time
as the Jython scripting interface,
and for that reason primarily
interacts with the simulator directly.
Since large-scale models often require
a scripting language for looping
and parameterization,
this has resulted in a difficult transition
for modellers moving from the
drag-and-drop graphical interface
to the scripting interface.
Additionally, advanced scripts often
contain Jython code that either defines
or manipulates Java objects directly,
meaning that modellers have to know
a certain amount of Python and Java to be productive.

Since Nengo makes it possible to define
very large models,
performance is an important issue.
While Java has good support for threads,
it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can quickly perform the types of parallel
computations that these neural models require.
This is due to the
architecture of the Java simulator
and the Java Native Interface,
which results in code that
is difficult to maintain and debug.

The new implementation of Nengo described in this paper
was undertaken with three goals in mind:
\begin{enumerate}
  \item Provide a scripting interface
    similar to the existing Jython interface.
  \item Leverage existing scientific computing tools
    rather than write our own.
  \item Provide the ability to run models with
    different software and hardware simulators.
\end{enumerate}

In order to achieve these goals,
this new version of Nengo uses the Python language
and targets the CPython interpreter,
allowing Nengo to use existing scientific Python
tools for vectorized computations, plotting, and graph manipulation.
As evidence that we have met these goals,
we have implemented two different software simulators,
and have done so with an order of magnitude
less lines of code than the Java simulator
and Jython scripting interface combined.
For the remainder of the paper,
we will use ``Nengo'' to refer to
this new Python rewrite.

\subsection{Neural Engineering Framework (NEF)}

>>>>

It is often said that neuroscience
is data-rich but theory-poor.
Although computational neuroscience
has grown significantly,
this phrase continues to be repeated
due to a lack of satisfying
theories of the brain.
Much work has been put into simulators
that attempt to recreate neural
data in precise detail,
but behavior has not emerged
from data-driven large scale models.
Similarly, theoretical frameworks
that involve a generic algorithm
replicated many times
(e.g., the memory-prediction framework \cite{TODO}
and the pattern recognition theory of mind \cite{TODO})
have not been shown to solve problems
outside of those that
the generic algorithm alone can solve.

Nengo is a neural simulator
based on a theoretical framework proposed
by Eliasmith \& Anderson \citeyearpar{TODO}
called the Neural Engineering Framework
(NEF).
Nengo was recently used to implement Spaun,
currently the world's largest functional brain model
\cite{TODO}.
Spaun is a network of 2.5 million spiking neurons
that can perform eight high-level cognitive tasks
including list memory and inductive reasoning.
It can perform any of these eight tasks
at any time by being presented
the appropriate series of images
representing the task to be performed
and the details of that task
(e.g., $A3[1234]$ instructs Spaun
to remember the list $1234$).
It then provides its answer by
sending motor commands to a simulated arm model,
tracing out digits
(e.g., $1234$ if asked to recall the remembered list).

While the tasks Spaun performs are diverse,
all tasks use a common set of
functional modules that we map
onto the brain areas hypothesized
to perform those functions.
As a result, Spaun is able to match
experimental data across the spectrum
of neuroscientific data,
from the single cell level
to the behavioral level,
despite experimental data
being a very small part of the model description.
Similarly, while this is made possible
by a theoretical framework,
the NEF enables Spaun to employ
many different algorithms with spiking neurons,
rather than defining a single algorithm
that is replicated many times.

In implementing Spaun,
we have provided strong support
that the NEF is beginning to fill the theory void in neuroscience.
However, we have also pushed Nengo
to the limits of its ability to construct and simulate
large-scale neural models.
In order to continue to evaluate the NEF
as a theory of the brain
by building larger and more complicated models,
we have begun the next generation
of Nengo tools with a focus on simplicity,
extensibility, and speed.

\subsection{JavaNengo limitations}

\begin{table}[!t]
\processtable{Nengo tools\label{tab:nengo-vers}}
{\begin{tabular}{p{2.6cm} p{2.7cm} l p{7cm}} \toprule
\textbf{Name} & \textbf{Implementation language}
  & \textbf{Code repository} & \textbf{Purpose} \\\midrule
JavaNengo simulator & Java & \texttt{nengo\_java}
  & Simulates many types of spiking neuron models.
  Includes methods to construct and simulate models using the NEF. \\
JavaNengo GUI & Java & \texttt{nengo\_java\_gui}
  & Graphical interface to JavaNengo simulator.
  Allows modellers to drag-and-drop components built with the NEF. \\
JavaNengo scripting interface & Jython & \texttt{nengo\_jython}
  & Scripting interface to JavaNengo simulator.
  Provides shortcuts to classes and functions in the simulator.
  Also accessible through and interacts with the JavaNengo GUI. \\
PyNengo & Python & \texttt{nengo}
  & Constructs and simulates models using the NEF. \\
PyNengo OpenCL simulator & Python+PyOpenCL
  & \texttt{nengo\_ocl}
  & Simulates models constructed with PyNengo. \\\botrule
\end{tabular}}{}
\end{table}

JavaNengo is a suite of tools providing
a neural simulator and graphical user interface (GUI)
implemented in Java,
and a scripting interface implemented in Jython
(see Table~\ref{tab:nengo-vers} for more detail).
JavaNengo's simulator was created
with the intention of being a general
neural simulator that included
the methods of the NEF
as one of many different
methods of creating neural models.
As a result, its implementation
of important NEF components
are nested in deep class hierarchies,
adding unnecessary layers of complexity
for developers wishing
to extend those NEF components.
The addition of synaptic plasticity rules,
for example, involved adding or significantly modifying
25 Java classes \cite{TODO}.\footnote{Changelog at
  \url{https://github.com/ctn-waterloo/nengo_1.4/commit/93eee51147545f7ce30ae9919809e3072dc15728}.}

The JavaNengo GUI provided
a drag-and-drop interface in which to create NEF models.
However, as it was developed
before the scripting interface,
the GUI manipulates simulator objects directly.
Typically, modellers start by following
tutorials that use the GUI interface,
and transition to the scripting interface
in order to implement more complicated models.
Since the GUI modifies simulator objects directly,
modelling concepts learned in the GUI
have to be relearned in the scripting interface.

The JavaNengo scripting interface,
described in \cite{TODO},
provided a simpler interface
to the NEF-specific portions of the simulator.
This greatly enhanced productivity,
and made it possible
to create many models with very little code.
Modellers quickly started using
the scripting interface
as their primary tool,
but still needed to access the GUI
for interactive model inspection,
and still needed to modify the simulator
to add new functionality
(e.g., new neuron models).
This is because Jython makes it easy
to inspect and create Java objects,
but does not make it easy to
write Java code to
inspect and create Jython objects.
Modellers, therefore, had to be
proficient in Java whether
they used the scripting interface or not.
As the set of neural modellers
that are proficient with both Java and Python
syntax is small,
the scripting interface still lacks
convenient ways to record the results of simulations,
and to define complex experimental setups.

Finally, perhaps the most significant limitation
of the JavaNengo tools is its performance.
While Java has good support for threads,
it does not have well-supported libraries
for doing mathematical operations on large matrices quickly.
Additionally, it is difficult to interface with
platform-specific tools such as
graphical processing units (GPUs),
which can perform the types of computations
that these neural models require quickly and in parallel.

Addressing all of these issues
in the Java codebase would involve
significant enough changes
that we decided to start from scratch instead.
Since the scripting interface
already used Python syntax,
and we were using Python
and its scientific libraries
for analyzing simulation results,
we targeted the CPython interpreter
as the platform for the new generation of Nengo tools.

\subsection{PyNengo}

PyNengo, which we describe in more detail
in the remainder of this paper,
is a Python package for
defining and simulating
neural models using the NEF.
It targets the CPython interpreter,
meaning that it integrates seamlessly
with other scientific Python tools.
Nengo models can used
in arbitrary CPython programs,
opening up possibilities
for using neurally inspired algorithms
in applications and games.

PyNengo's object model has been radically simplified,
making it easy to
document, test, and modify those objects.
It also includes a scripting interface to those objects
based on JavaNengo's scripting interface \cite{TODO},
but with several modifications
to allow for easier incorporation
of new neuron models and learning rules.
Additionally, PyNengo's scripting interface
contains methods to record simulated data
and access that data after simulation.

Model creation and simulation
have been completely decoupled,
allowing for models to be exported
to other simulators,
such as those accessible with PyNN.
Simulators can also be written
to leverage specific platforms.
Currently, we have implemented
a platform-independent simulator
that uses NumPy for vectorized computations,
and a PyOpenCL-based
simulator that takes full advantage of GPUs
and multicore CPUs.

PyNengo does not currently
replace all of the JavaNengo tools.
Most notably, there is no graphical interface
to PyNengo, and therefore little ability to interact
with simulations as they are running.
However, we are confident that these tools
can be created, and will be much easier
to implement compared to JavaNengo's GUI.

We believe that PyNengo overcomes
all of the significant issues
with JavaNengo.
It can simulate the majority of models
currently simulated by JavaNengo,
using 11\% as many lines code,
and depending on
only three external libraries instead of ten.
The OpenCL simulator can simulate
a large scale network TODO times
faster than JavaNengo's simulator.
This, along with a codebase
that is easier to understand and extend,
provides a platform
that can simulate larger and more complex
models than Spaun,
and can therefore further
test the NEF as a theory of the brain.

\section{Neural Engineering Framework (NEF)}
